---
title: "Trajectory Processing"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
`%>%` <- magrittr::`%>%`
library(tidyverse)
library(plotly)
```

A set of test data has been downloaded from Opensky Network. 
With a view to produce the analytic data, the following steps are required:

* read in trajectory, clean spikes, and complement missing items;
* apply heuristics to define key movement timestamps, e.g. AOBT, ATOT;
* trajectory simplification (maybe before - tbd --> Schultz et al., 2020/2019)

## Check Coverage

For comparison reasons, we read in data from the APDF to check on coverage and movement association.

```{r}
study_apt  <- "LSZH"
check_apdf <- paste0("./data/", study_apt, "_APDF_CHECK.csv")
col_specs <- readr::cols_only(
   AP_C_FLTID = readr::col_character()
  ,AP_C_REG   = readr::col_character()
  ,ADEP_ICAO  = readr::col_character()
  ,ADES_ICAO  = readr::col_character()
  ,AP_C_RWY   = readr::col_character()
  ,AP_C_STND  = readr::col_character()
  ,MVT_TIME_UTC  = readr::col_character()
  ,BLOCK_TIME_UTC= readr::col_character()
  ,SCHED_TIME_UTC= readr::col_character()
  ,ARCTYP     = readr::col_character()
  ,AC_CLASS   = readr::col_character()
  ,SRC_PHASE  = readr::col_character()
)
### default col_character: lubridate::parse_date_time(AIBT, orders = c("Ymd HMS","dmY HMS")
if(!file.exists(check_apdf)){
  message(paste0("Reading in APDF Fact Table for ", study_apt))
  fn   <- paste0("../__DATA/", study_apt,"/", study_apt, "_2019_FACT.csv") 
  apdf <- readr::read_csv2(fn, col_types = col_specs) %>%
    mutate(
      MVT_TIME_UTC = lubridate::parse_date_time(MVT_TIME_UTC, orders = c("dmY HMS"))
   ,BLOCK_TIME_UTC = lubridate::parse_date_time(BLOCK_TIME_UTC, orders = c("dmY HMS"))
   ,SCHED_TIME_UTC = lubridate::parse_date_time(SCHED_TIME_UTC, orders = c("dmy HMS"))
    )
  readr::write_csv(apdf, check_apdf)
}else{
  message(paste0("loading APDF reference for ", study_apt))
  apdf <- readr::read_csv(check_apdf, col_types = readr::cols(.default = "c")) %>%
    dplyr::mutate(
      MVT_TIME_UTC = lubridate::parse_date_time(MVT_TIME_UTC, orders = c("Ymd HMS"))
   ,BLOCK_TIME_UTC = lubridate::parse_date_time(BLOCK_TIME_UTC, orders = c("Ymd HMS"))
   ,SCHED_TIME_UTC = lubridate::parse_date_time(SCHED_TIME_UTC, orders = c("Ymd HMS"))
      )
}
```

For the initial development we are working with the data for 1. October.

```{r}
start_date = as.POSIXct("2019-10-01 00:00:00", tz="UTC")
end_date  <- start_date + lubridate::days(1)

apdf2 <- apdf %>% 
  dplyr::filter(BLOCK_TIME_UTC >= start_date & BLOCK_TIME_UTC < end_date)
```

Let's do a quick visual check with ggplot for traffic binned to UTC hours on the 1. October 2019.

```{r}
tfc_hr <- apdf2 %>% 
  mutate(BIN = lubridate::floor_date(BLOCK_TIME_UTC, unit = "hour")
         ,HR = lubridate::hour(BLOCK_TIME_UTC)) %>%
  group_by(SRC_PHASE, BIN, HR) %>%
  summarise(FLIGHTS = n())

g <- ggplot(data = tfc_hr, mapping = aes(x = BIN, y = FLIGHTS, colour = SRC_PHASE)) +
  geom_step() +
  theme_minimal()
ggplotly(g) %>%
  layout(hovermode = "x")
#g
```

There is a departure peak (AOBT) between 0900 and 1200hrs UTC.

```{r}
read_data <- function(.pattern, .path, .df=TRUE, .cols){
  data_files <- list.files(path = .path, pattern = .pattern )
  file_names <- paste0(.path,"/", data_files)
  
  if(.df){
    ds <- file_names %>% purrr::map_dfr(.f=read_csv, col_types = .cols)
  }else{
    ds <- file_names %>% purrr::map(.f=read_csv)
  }
  return(ds)
}

col_specs <- cols(
  time = col_double(),
  icao24 = col_character(),
  lat = col_double(),
  lon = col_double(),
  velocity = col_double(),
  heading = col_double(),
  vertrate = col_double(),
  callsign = col_character(),
  onground = col_logical(),
  alert = col_logical(),
  spi = col_logical(),
  squawk = col_double(),
  baroaltitude = col_double(),
  geoaltitude = col_double(),
  lastposupdate = col_double(),
  lastcontact = col_double(),
  hour = col_double()
)

ds <- read_data(.pattern = "osn_LSZH_R20_", .path = "./data-src", .cols=col_specs) 
```

Opensky Network data expresses the timestamps in Unix epochs.

```{r}
ds <- ds %>% mutate(time = lubridate::as_datetime(time, tz="UTC"))
```

Let's pick some test trajectories during the peak period (0900 and 1200hrs UTC).
For this we allow read-outs 5 minutes prior and after the peak period.

```{r}
pick_start <- as.POSIXct("2019-10-01 08:55:00", tz = "UTC")
pick_end   <- as.POSIXct("2019-10-01 12:05:00", tz = "UTC")

pick <- ds %>% 
  filter(time >= pick_start, time <= pick_end)

# summary stats
pick_stats <- pick %>% 
  group_by(icao24) %>%
  summarise(HITS = n(), ONG = sum(onground, na.rm = TRUE), ONG_P = ONG / HITS)
```

From the aforementioned APDF data feed we extract random departures during the peak horizon.

```{r}
some_deps <- apdf2 %>%
  filter(ADEP_ICAO == "LSZH", BLOCK_TIME_UTC >= pick_start, BLOCK_TIME_UTC <= pick_end)
```

> some_deps %>% arrange(BLOCK_TIME_UTC) %>% sample_n(5) %>% select(FLTID = AP_C_FLTID, REG = AP_C_REG, ADEP = ADEP_ICAO, ADES = ADES_ICAO, AOBT = BLOCK_TIME_UTC, ATOT = MVT_TIME_UTC)
# A tibble: 5 x 6
  FLTID   REG   ADEP  ADES  AOBT                ATOT               
  <chr>   <chr> <chr> <chr> <dttm>              <dttm>             
1 DLH1187 DAIWH LSZH  EDDF  2019-10-01 09:25:15 2019-10-01 09:34:35
2 BAW711  GEUXE LSZH  EGLL  2019-10-01 09:38:00 2019-10-01 09:52:00
3 ATL3Z   DINOB LSZH  EBKT  2019-10-01 09:54:39 2019-10-01 09:58:16
4 SWR1662 HBIJN LSZH  LIPZ  2019-10-01 11:21:00 2019-10-01 11:37:15
5 VLG8349 ECNEA LSZH  LEPA  2019-10-01 11:33:46 2019-10-01 11:52:38

```{r}
fltids <- c("DLH1187","BAW711","ATL3Z","SWR1662","VLG8349")

trjs <- ds %>% select(icao24, callsign) %>% filter(callsign %in% fltids) %>%
  unique()
trjs
```

```{r}
trjs <- ds %>% filter(icao24 %in% c("400afd", "346185"))

g <- ggplot() + 
  geom_point(data = trjs, mapping = aes(x = lon, y = lat, colour = callsign))+
  theme_minimal() +
  coord_map()
ggplotly(g)
```

Vertical Profile

```{r}
plot_ly(data = trjs) %>%
  add_lines(x = ~time, y = ~ geoaltitude, color = ~callsign)
```

The vertical profile shows spikes for the altitude readout of both flights.
The departure phase seems to be well tracked (without spikes).
It follows that the callsign field can be used to separate successive flights of an aircraft (= same ICAO-24bit address).    
Note: While a callsign change is indicative for successive flights, there are more characteristics. For example, aircraft remains on the ground (for a sufficient time). Touch-and-go or low approaches can be detected based on same icao24, virtually no ground time, and then departing leg.

Let's analyse the trajectory 

```{r}
#' Estimate airport flight phase
#'
#' Utility function to estimate the flight phase (i.e. arrival or departure)

est_apt_flt_phase <- function(.trjs){
  df <- .trjs
  df <- df %>% group_by(icao24, callsign) %>% 
    summarise(
       HITS = n()
      ,ONG  = sum(onground, na.rm = TRUE)
      ,FIRST= min(time)
      ,FIRST_ALT = geoaltitude[time == FIRST]
      ,LAST = max(time)
      ,LAST_ALT  = geoaltitude[time == LAST]
      ) %>%
    mutate(
       DUR_COV = difftime(LAST, FIRST, units = "mins")
      ,ALT_CHG = case_when(
         is.na(FIRST_ALT) & LAST_ALT  >= 500  ~ "DEP"
        ,is.na(LAST_ALT)  & FIRST_ALT >= 500  ~ "ARR"
        ,FIRST_ALT >= 3000 & LAST_ALT >= 3000 ~ "OVR"
        ,TRUE ~ "DUNNO"
      ) 
    )
  return(df)
}

trjs %>% est_apt_flt_phase()
```

Note: TO-DO ... need to clean trajectories for NAs and altitude.

Apply heuristic to dataset:

```{r}
flt_table <- ds %>% est_apt_flt_phase()
flt_table
```

Special DUNNO cases --> FXT401, MMD6449  ==> multi-leg flights (same icao24 and same callsign)

```{r}
tmp <- ds %>% filter(callsign %in% c("FXT401","MMD6449","AHO253Q","VJT486","NJE720G","PEA301","HBZUR","DUKE15","PEG81"))
plot_ly(data = tmp) %>%
  add_lines(x = ~time, y = ~ geoaltitude, color = ~callsign)
```

Exclude obvious non-departures or non-arrivals

think about this more careful

```{}
#' Identify airport traffic
#'
#' Utility function to remove non airport related traffic without using
#' distance

trim_apt_tfc <- function(.trjs){
  df <- .trjs
  # filter obvious non-arriving or non-departing traffic
  df <- df %>% 
  
}
```







