---
title: Data Preparation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(magrittr)
```

This page documents the data preparatory action.

## Download Airport Lookup Reference Table

Aiport reference data is downloaded from OpenAirports.com.

```{r}
# airport look-up table
apt_lu <- "./data/apt_lu.csv"

if(!file.exists(apt_lu)){
  message("Downloading airport reference table from OpenAirports.com.")
  url    <- "https://ourairports.com/data/airports.csv"
  apt_lu <- readr::read_csv(url)
  readr::write_csv(apt_lu, "./data/apt_lu.csv")
}else{
  message("Loading airport reference table from OpenAirports.com")
  apt_lu <- readr::read_csv("./data/apt_lu.csv")
}
```

Set-up the table of refernces for this study.

```{r}
study_apts <- apt_lu %>%
  dplyr::filter(ident %in% c("LSZH", "EGKK", "EDDF"))
```

* ident ICAO location indicator
* latitude_deg latitude of the aerodrome reference point
* longitude_deg longitude of the aerodrome reference point

## Download ADSB State-Vectors from Opensky Network

For this paper state vectors for the following airports are downloaded:

* LSZH Zurich

With a focus to the airport environment and the associated ICAO GANP KPIs we (try to) download data within a bounding box of 100NM.
The bounding_box function is documented in R/bounding_box.R.

```{r}
source("./R/bounding_box.R")

apt <- study_apts %>% dplyr::filter(ident == "LSZH") %>% 
  dplyr::select(
     ICAO = ident
    , LON = longitude_deg
    , LAT = latitude_deg
    ,ELEV = elevation_ft
    )

apt_bb <- bounding_box(lat = apt$LAT, lon = apt$LON, dist = 20)
# coerce apt_bb matrix into vector c(LONmin, LONmax, LATmin, LATmax)
apt_bb <- apt_bb %>% t() %>% as.vector()
# apply naming convention for Opensky Network
names(apt_bb) <- c("xmin","xmax","ymin","ymax")
```

Establish session with Opensky Network Server

```{r}
source("./R/osn_query_apt_bbox.R")

# account details stored in environment
rq_usr <- "espin"     #Sys.getenv("osn_usr")
rq_pw  <- "Losvce1300"  #Sys.getenv("osn_pw")

# establish session with Opensky Network
session <- osn::osn_connect(usr = rq_usr, passwd = rq_pw)

start_date <- lubridate::ymd_hms("2019_10_01 09:00:00", tz = "UTC")
df <- query_osn(start_date, session, apt_bb)
```

c(xmin = 7.536746, xmax = 9.604390, ymin = 49.36732, ymax = 50.69920)